---
  title: "Convert hexadecimal files"
  output:
    md_document:
    variant: gfm
---
  
# Convert hexadecimal files to cnv

The gapctd package can convert Sea-Bird CTD hexadecimal files (.hex) to .cnv file with variables in scientific units. Currently, `hex_to_cnv()` supports SBE19plus temperature, conductivity, pressure, and flags, along with data from SBE43 dissolved oxygen and SBE18 pH sensors. The default outputs are:

- Temperature [ITS-90, degrees C]
- Conductivity [S/m]
- Pressure [decibars]
- Data quality flags
- Dissolved oxygen [voltage]
- pH [voltage]

The default output does not include salinity, oxygen, or other derived quantities because additional processing is required to correct for sensor lag times and dynamic errors.

This code below uses hexadecimal and configuration (.xmlcon) files as inputs to generate a .cnv file named 2021_06_24_0001.cnv. The data are then loaded as `ctd` objects using the oce package.


#### Conversion
```{r hex_to_cnv, warnings=FALSE, message=FALSE, include=TRUE, echo=TRUE, eval=TRUE}
library(gapctd)

hex_to_cnv(hex_path = system.file("extdata/example/2021_06_24_0001.hex", package = "gapctd"),
           xmlcon_path = system.file("extdata/example/19-8102_Deploy2021.xmlcon", package = "gapctd"),
           output_path = "2021_06_24_0001.cnv")
```

#### Loading as ctd object

```{r load_cnv, warnings=FALSE, message=FALSE, include=TRUE, echo=TRUE, eval=TRUE}

dat <- oce::read.oce(file = "2021_06_24_0001.cnv")

plot(dat)
```