---
  title: "Convert hexadecimal (.hex) files to .cnv"
  output:
    md_document:
    variant: gfm
---
  
# Convert hexadecimal files to cnv

The gapctd package can convert Sea-Bird CTD hexadecimal files (.hex) to .cnv file with variables in scientific units. Currently, `hex_to_cnv()` supports SBE19plus temperature, conductivity, pressure, and flags, along with data from SBE43 dissolved oxygen and SBE18 pH sensors. The default outputs are:

- Temperature [ITS-90, degrees C]
- Conductivity [S/m]
- Pressure [decibars]
- Data quality flags
- Dissolved oxygen [voltage]
- pH

Note that the cnv file created by `hex_to_cnv()` does not include derived quantities that depend on dynamic error corrections from primary channels (e.g. salinity, density, dissolved oxygen). The `oce::read.ctd()` package automatically calculates PSS-78 when it is not provided, although this value should not be used directly. Dynamic errors should first be corrected (see [Data processing modules](./doc/data_processing_modules.md), [Batch processing steps](./doc/batch_processing_steps.md), and [Batch processing instructions and example](./doc/0_batch_processing.md))

This code below uses hexadecimal and configuration (.xmlcon) files as inputs to generate a .cnv file named 2021_06_24_0001.cnv. The data are then loaded as `ctd` objects using the oce package.


#### Converting data without auxillary sensors

```{r hex_to_cnv, warnings=FALSE, message=FALSE, include=TRUE, echo=TRUE, eval=TRUE}
library(gapctd)

hex_to_cnv(hex_path = system.file("extdata/example/2021_06_24_0001.hex", package = "gapctd"),
           xmlcon_path = system.file("extdata/example/19-8102_Deploy2021.xmlcon", package = "gapctd"),
           output_path = "2021_06_24_0001.cnv")

dat <- oce::read.oce(file = "2021_06_24_0001.cnv")

plot(dat)
```

#### Converting data with dissolved oxygen and pH sensors

```{r do_ph, warnings=FALSE, message=FALSE, include=TRUE, echo=TRUE, eval=TRUE}
library(gapctd)

hex_to_cnv(hex_path = system.file("extdata/example/SBE19plus_01908106_2023_06_19_0001.hex", package = "gapctd"),
           xmlcon_path = system.file("extdata/example/SBE19plusV2_8106_ph_DO_leg2.xmlcon", package = "gapctd"),
           output_path = "SBE19plus_01908106_2023_06_19_0001.cnv")

dat_doph <- oce::read.oce(file = "SBE19plus_01908106_2023_06_19_0001.cnv")

plot(dat_doph)

head(as.data.frame(dat_doph@data))

```