---
title: "Data Processing Modules in R"
output:
  md_document:
    variant: gfm
---

The gapctd package includes modules (filters and operations) for processing CTD data that are based on modules in SBE Data Processing (SBEDP) software. The gapctd modules replicate the output of SBEDP modules. However, the gapctd modules do not always include the full range of processing options that are available in SBEDEP modules because the gapctd modules are intended for limited use cases.

Modules accept and return oce objects and are designed to be interchangable (i.e., the order in which they are used can be swapped). This document describes the modules and demonstrates their use for processing CTD data.

## Load example data

Read-in the example CTD deployment file (.cnv) as a `ctd` object using the `read.oce` function from the oce package. The deployment file contains five variable columns that are populated in the oce object. A sixth field, salinity (simply called 'salinity'), is derived when the data are read-in using `read.oce`.

| CTD File (.cnv) Variable | `ctd` object Variable | Description |
|------|------|------|
| timeS | timeS | Time, Elapsed [seconds] |
| tv290C | temperature | Temperature [ITS-90, deg C] |
| prdM | pressure | Pressure, Strain Gauge [db] |
| c0S/m | conductivity | Conductivity [S/m] |
| flag | flag | Data quality flag |
| - |  salinity | Practical Salinity [PSS-78] |

The file includes data from the entire deployment but only the downcast will be used for this demonstration. The `oce::ctdTrim` function is used to automatically detect and remove scan data that were not in the downcast.

```{r load_libraries, warnings=FALSE, message=FALSE, include=FALSE, echo=FALSE, eval=TRUE}
library(gapctd)

ctd_data <- oce::read.oce(file = system.file("extdata/example/2021_06_24_0001_raw.cnv", package = "gapctd"))
dc <- oce::ctdTrim(ctd_data, method = "sbe")
```

```{r load_libraries_2, include=TRUE, eval=FALSE}
library(gapctd)

ctd_data <- oce::read.oce(file = system.file("extdata/example/2021_06_24_0001_raw.cnv", package = "gapctd"))
dc <- oce::ctdTrim(ctd_data, method = "sbe")

plot(dc)
```


## Modules

### Median window filter

The `median_filter` module is based on the median filter option in the SBEDP Window Filter (Wfilter) module. It calculates the median for selected channels within a specified scan window. Below, the median window filter is applied to temperature and conductivity channels for five (5) scan windows (corresponding with 1.25 seconds for the SBE19plus V2) and salinity is recalculated.


```{r}
dc_1 <- gapctd::median_filter(dc, 
                              variables = c("temperature", "conductivity"),
                              window = c(5,5))
dc_1@data$salinity <- oce::swSCTp(dc_1) # Calculate salinity

par(mfrow = c(1,2))
plot(dc, which = 1, type = 'l')
text(x = 31.56, y = 35, labels = "(1) Raw Downcast")
plot(dc_1, which = 1, type = 'l')
text(x = 31.56, y = 35, labels = "(2) Median Filter\nT+C")
```


### Lowpass filter

The `lowpass_filter` module is based on the SBEDP Filter module. Below, the filter is applied to temperature, conductivity, and pressure channels with time constants of 0.5 s, 0.5 s, and 1 s, respectively. Salinity is then recalculated.


```{r}
dc_2 <-   gapctd::lowpass_filter(dc_1,
                                 variables = c("temperature", "conductivity", "pressure"),
                                 time_constant = c(0.5, 0.5, 1))
dc_2@data$salinity <- oce::swSCTp(dc_2) # Calculate salinity

par(mfrow = c(1,2))
plot(dc_1, which = 1, type = 'l')
text(x = 31.56, y = 35, labels = "(2) Median Filter\nT+C")
plot(dc_2, which = 1, type = 'l')
text(x = 31.56, y = 35, labels = "(3) Low pass filter")
```



### Align channel

The `align_var` module is based on the SBEDP Align (AlignCTD) module. The module shifts channel data in time and interpolates between measurements when alignment parameters are not factors of the scan interval. In the example below, temperature is advanced by 0.5 seconds and salinity is recalculated. Plots showing a close-up of the section between 10 and 12 dbar.

```{r}
dc_3 <-   gapctd::align_var(dc_2,
                            variables = "temperature", 
                            offset = -0.5)
dc_3@data$salinity <- oce::swSCTp(dc_3) # Calculate salinity

par(mfrow = c(1,2))
plot(dc_2, which = 1, type = 'l', plim = c(12,10))
text(x = 31.55, y = 11.85, labels = "(3) Low pass filter")
plot(dc_3, which = 1, type = 'l', plim = c(12,10))
text(x = 31.55, y = 11.85, labels = "(4) Align T")
```


### Conductivity cell thermal mass correction

The `conductivity_correction` module is based on the SBEDP Conductivity Cell Thermal Mass Correction (CellTM) module. The module implements a discrete time filter on conductivity based on the sample interval (freq_n), the initial conductivity error (alpha_C), and a time decay constant (beta_C). The function uses the sample interval in the data to estimate the sample interval if it is not provided.

Below, the conductivity correction is applied using the typical parameters recommended by the CTD manufacturer and salinity is recalculated.

```{r}
dc_4 <- gapctd::conductivity_correction(dc_3,
                                        alpha_C = 0.04, 
                                        beta_C = 1/8,
                                        freq_n  = 0.25)
dc_4@data$salinity <- oce::swSCTp(dc_4)

par(mfrow = c(1,2))
plot(dc_3, which = 1, type = 'l')
text(x = 31.58, y = 35, labels = "(4) Align T")
plot(dc_4, which = 1, type = 'l')
text(x = 31.58, y = 35, labels = "(5) Correct C")
```


### Loop edit

The `loop_edit` module is based on the SBEDP Loop Edit (loopedit) module that flags slowdowns and reversals in scan data based on a threshold. 

```{r}
dc_5 <- gapctd::loop_edit(dc_4,
                          min_speed = 0.1, 
                          window = 5, 
                          cast_direction = "downcast")

dc_4@data$flag[409:415]
dc_5@data$flag[409:415]

par(mfrow = c(1,2))
plot(dc_4, which = 1, type = 'l')
text(x = 31.58, y = 35, labels = "(5) Correct C")
plot(dc_5, which = 1, type = 'l')
text(x = 31.58, y = 35, labels = "(6) Loop edit")
```

Note that the module flags slowdowns and reversals but does not remove them, so the plots above are identical.

### Derive EOS

The `derive_eos` module uses functions from the oce package to calculate the following variables from scan data:

| Variable | Description | Method | `oce` function |
|------|------|------|------|
| depth | Depth [m] | EOS-80 | swDepth() |
| salinity | Practical salinity [PSS-78] | EOS-80 | swSCTp(eos="unesco") |
| density | In-situ density [kg m^-3] | EOS-80 | swRho() |
| absolute_salinity | Absolute salinity | TEOS-10 | swSCTp(eos="gsw") |
| N2 | Square of buoyancy frequency [radians^2 s^-2] | | swN2() |

```{r}
dc_6 <- gapctd::derive_eos(dc_5)

head(as.data.frame(dc_6@data))
```


### Bin average

The `bin_average` module is based on the SBDEP Bin Average (BinAvg) module. The module calculates means of variables by depth or pressure bin, with bad scan (flag < 0) data excluded. Binning can be performed by pressure or depth bins and surface data can be excluded based on a user-specified depth or pressure threshold (default = 0.5). In the example below, means are calculated for 1 m depth bins and data from < 0.5 m depth are excluded.

```{r}
dc_7 <- gapctd::bin_average(dc_6, by = "depth", bin_width = 1)

par(mfrow = c(1,2))
plot(dc_5, which = 1, type = 'l')
text(x = 31.58, y = 35, labels = "(6) Loop edit")
plot(dc_7, which = 1, type = 'l')
text(x = 31.58, y = 35, labels = "(7) Bin average")
```


### Linking modules with pipe operators

To declutter the workspace, the entire workflow can be run using pipe operators (native `|>` or magrittr `%>%` instead of producing output at every step. This is because each module accepts and returns a `ctd` object.

```{r}
dc_8 <- dc |>
  gapctd::median_filter(variables = c("temperature", "conductivity"),
                        window = c(5,5)) |>
  gapctd::lowpass_filter(variables = c("temperature", "conductivity", "pressure"),
                         time_constant = c(0.5, 0.5, 1)) |>
  gapctd::align_var(variables = "temperature", 
                    offset = -0.5) |>
  gapctd::conductivity_correction(alpha_C = 0.04, 
                                  beta_C = 1/8) |>
  gapctd::loop_edit(min_speed = 0.1, 
                    window = 5, 
                    cast_direction = "downcast") |>
  gapctd::derive_eos() |>
  gapctd::bin_average(by = "depth", bin_width = 1)

plot(dc_7, which = 1, type = 'l')
text(x = 31.58, y = 35, labels = "(8) Binned")
plot(dc_8, which = 1, type = 'l')
text(x = 31.58, y = 35, labels = "(9) Piped")
```


### Comparison with gapctd output

The processing steps above are similar to the process that is implemented in the package when processing using the full gapctd workflow. The plots below show the results after bin average and the same profile processed using the full processing method in the gapctd package.

```{r}
final_data <- readRDS(file = system.file("extdata/example/2021_06_24_0001_final.rds", package = "gapctd"))

par(mfrow=c(1,2))
plot(dc_7, which = 1, type = 'l')
text(x = 31.58, y = 35, labels = "(7) Binned")
plot(final_data$downcast, which = 1, type = 'l')
text(x = 31.57, y = 33, labels = "Processed w/\ngapctd")
```


### Modules for workign with GAP data

The modules above can be used with any `ctd` object data from any CTD. However, there are some modules that have specifically been developed for processing GAP's CTD data:

| Module | Description |
|------|------|
| `append_hau_data()`| |




