---
title: "Optimizing Cell Thermal Mass"
output:
  md_document:
    variant: markdown_github
---

##  Optimizing Cell Thermal Mass parameters

Conductivity cell thermal mass corrections correct for lags in conductivity measurements due to the thermal inertia of the conductivity cell. Conductivity measurements are temperature dependent but it takes time for the conductivity cell to heat and cool as it passes through temperature gradients. Because salinity is calculated from conductivity and temperature, the lag in conductivity can cause erroneous spikes in salinity around temperature gradients. The spikes are typically negative (i.e., toward lower salinity) as the CTD passes from warm to cold, and positive when passing from cold to warm. As such, upcasts and downcasts in a thermally stratified water column tend to have spikes in opposite directions around large temperature gradients.

Cell thermal mass corrections are performed using a recursive filter (Lueck and Pickolo, 1990) that corrects conductivity measurements sequentially based on changes in temperature between scans and the scan frequency (4 Hz for bottom trawl survey data collected using an SBE19plus). For the SBE19plus, the manufacturer recommends 'typical' cell thermal mass correction parameters &alpha; = 0.04 and &tau; = 8. However, the optimal parameters for cell thermal mass correction can differ from the optimal values depending on the magnitude of initial conductivity errors that can vary depending on the steepness of temperature gradients and instrument profiling rate (Mensah et al., 2009; Martini et al., 2019). Rather than use a single set of thermal mass correction parameters, the gapctd package uses the `ctm_adjust_ts_area` function to estimate cell thermal mass correction parameters (&alpha; and &tau;) for each deployment based on the method of Garau et al. (2011), who estimated parameters by minimizing the area between temperature-salinity curves from autonomous glider downcasts and upcasts.

## Load packages and data

Load the `gapctd` package and example upcast and downcast data from the same haul in the eastern Bering Sea. these data were processed using the [sbe19plus_v0 method](batch_processing_methods.md) then aligned using [estimated temperature adjustments](estimating_alignment.md) (downcast +0.73 seconds, upcast +0.77 seconds).

```{r setup, include=FALSE, warning=FALSE, results ='hide'}
library(gapctd)
library(ggthemes)

align_down_oce <- oce::read.oce(system.file("./extdata/ex_data/ctm_downcast_ex.cnv", package = "gapctd"))
align_up_oce <- oce::read.oce(system.file("./extdata/ex_data/ctm_upcast_ex.cnv", package = "gapctd"))

no_ctm_df <- dplyr::bind_rows(data.frame(timeS = align_down_oce@data$timeS[align_down_oce@data$pressure > 4],
  temperature = align_down_oce@data$temperature[align_down_oce@data$pressure > 4], 
                                         pressure = align_down_oce@data$pressure[align_down_oce@data$pressure > 4],
                                         conductivity = align_down_oce@data$conductivity[align_down_oce@data$pressure > 4],
                                         salinity = align_down_oce@data$salinity[align_down_oce@data$pressure > 4],
                                         flag = align_down_oce@data$flag[align_down_oce@data$pressure > 4],
                                         direction = "down"),
                              data.frame(timeS = align_up_oce@data$timeS[align_up_oce@data$pressure > 4],
                                         temperature = align_up_oce@data$temperature[align_up_oce@data$pressure > 4], 
                                         pressure = align_up_oce@data$pressure[align_up_oce@data$pressure > 4],
                                         conductivity = align_up_oce@data$conductivity[align_up_oce@data$pressure > 4],
                                         salinity = align_up_oce@data$salinity[align_up_oce@data$pressure > 4],
                                         flag = align_up_oce@data$flag[align_up_oce@data$pressure > 4],
                                         direction = "up"))

no_ctm_binned_df <- no_ctm_df |>
  dplyr::filter(pressure > 4, !is.na(flag)) |>
  dplyr::mutate(pressure_bin = round(pressure)) |>
  dplyr::group_by(direction, pressure_bin) |>
  dplyr::summarise(salinity = mean(salinity),
                   temperature = mean(temperature))
```

```{r load, include=TRUE, eval=FALSE}
library(gapctd)
library(ggthemes)

align_down_oce <- oce::read.oce("./inst/extdata/ex_data/ctm_downcast_ex.cnv")
align_up_oce <- oce::read.oce("./inst/extdata/ex_data/ctm_upcast_ex.cnv")

no_ctm_df <- dplyr::bind_rows(data.frame(timeS = align_down_oce@data$timeS[align_down_oce@data$pressure > 4],
  temperature = align_down_oce@data$temperature[align_down_oce@data$pressure > 4], 
                                         pressure = align_down_oce@data$pressure[align_down_oce@data$pressure > 4],
                                         conductivity = align_down_oce@data$conductivity[align_down_oce@data$pressure > 4],
                                         salinity = align_down_oce@data$salinity[align_down_oce@data$pressure > 4],
                                         flag = align_down_oce@data$flag[align_down_oce@data$pressure > 4],
                                         direction = "down"),
                              data.frame(timeS = align_up_oce@data$timeS[align_up_oce@data$pressure > 4],
                                         temperature = align_up_oce@data$temperature[align_up_oce@data$pressure > 4], 
                                         pressure = align_up_oce@data$pressure[align_up_oce@data$pressure > 4],
                                         conductivity = align_up_oce@data$conductivity[align_up_oce@data$pressure > 4],
                                         salinity = align_up_oce@data$salinity[align_up_oce@data$pressure > 4],
                                         flag = align_up_oce@data$flag[align_up_oce@data$pressure > 4],
                                         direction = "up"))

no_ctm_binned_df <- no_ctm_df |>
  dplyr::filter(pressure > 4, !is.na(flag)) |>
  dplyr::mutate(pressure_bin = round(pressure)) |>
  dplyr::group_by(direction, pressure_bin) |>
  dplyr::summarise(salinity = mean(salinity),
                   temperature = mean(temperature), 
                   .groups = "keep")
```

Profiles (Figure 1) show a two-layer stratified water column. Temperature and conductivity appear to be aligned, but downcast and upcast profiles have salinity spikes in the opposite directions around the pycnocline. As such a conductivity cell thermal mass correction is warranted.

```{r, echo = FALSE}
ggplot() +
  geom_path(data = no_ctm_df |>
              dplyr::filter(pressure > 4, !is.na(flag)) |>
              dplyr::mutate(pressure_bin = round(pressure)) |>
              dplyr::group_by(direction, pressure_bin) |>
              dplyr::summarise(salinity = mean(salinity),
                               temperature = mean(temperature),
                               conductivity = mean(conductivity)) |>
              tidyr::pivot_longer(cols = c("salinity", "temperature", "conductivity")),
            aes(x = value, y = pressure_bin, color = direction)) +
  scale_color_colorblind("Direction") +
  scale_y_reverse(name = "Pressure (db)") +
  facet_wrap(~factor(name, levels = "temperature", "conductivity", "salinity"), scales = "free_x")
```
<i>Figure 1. Temperature, conductivity, and salinity versus pressure from a CTD profiled collected during AFSC's 2021 eastern Bering Sea continental shelf bottom trawl survey.</i>

## Estimating CTM parameters

The gapctd package uses the `ctm_adjust_tsarea` function to estimate optimal parameters for cell thermal mass (CTM) correction by minimizing the area between temperature-salinity curves. The optimization starts at the manufacturer's recommended 'typical parameters (&alpha; = 0.04, &tau; = 8) and the scan interval (f_n) is set to 0.25 seconds for the 4 Hz sampling rate of the SBE18 plus. Scans with pressure < 4 db are not used for the optimization due to instances of erroneously low surface salinity and other artifacts while the instrument is being deployed. Numerical optimization is conducted using the Broyden-Fletcher-Goldfarb-Shanno algorithm (method = "BFGS") with a convergence tolerance of 1e-5 and maximum Newton steps = 500.

```{r, include=TRUE, echo=TRUE}
optim_results <- bbmle::mle2(minuslogl = gapctd:::ctm_adjust_tsarea,
                start = list(alpha = 0.04,
                             tau = 8),
                data = list(down_temperature = align_down_oce@data$temperature[align_down_oce@data$pressure > 4],
                            down_pressure = align_down_oce@data$pressure[align_down_oce@data$pressure > 4],
                            down_conductivity = align_down_oce@data$conductivity[align_down_oce@data$pressure > 4],
                            down_flag = align_down_oce@data$flag[align_down_oce@data$pressure > 4],
                            up_temperature = align_up_oce@data$temperature[align_up_oce@data$pressure > 4],
                            up_pressure = align_up_oce@data$pressure[align_up_oce@data$pressure > 4],
                            up_conductivity = align_up_oce@data$conductivity[align_up_oce@data$pressure > 4],
                            up_flag = align_up_oce@data$flag[align_up_oce@data$pressure > 4],
                            obj_fn = "area",
                            f_n = 0.25),
                method = "BFGS",
                control = list(maxit = 500, reltol = 1e-5, trace = 6,
                               parscale = c(alpha = 0.01, tau = 1)))
optim_results@coef
```
The algorithm converges and estimated parameters (&alpha; = -0.147, &tau; = 1.591) differ from the 'typical' starting values, which suggests the alternative parameters should be used. The package tries a few other starting values if the optimization does not converge and uses the manufacturer's typical values if none of the optimization attempts converge or if parameters estimates results in an optimized area between T-S curves that is higher than the area from the typical values. 

## Comparison with no CTM and CTM using typical parameters

Profiles that have the smallest area between T-S curves are eventually passed to later stages of processing and used to produce data products. It is therefore instructive to compare areas between T-S for downcast and upcast profiles without the correction (No CTM), CTM-corrected profiles using typical values (Default CTM), and CTM-corrected profiles using the optimal parameters (Optimized CTM).

```{r, include = TRUE, echo=TRUE}
ts_area_df <- data.frame(ctm = c("No CTM", "Default CTM", "Optimized CTM"),
           area = c(data.frame(temperature = c(no_ctm_binned_df$temperature[no_ctm_binned_df$direction == "down"],
                                               rev(no_ctm_binned_df$temperature[no_ctm_binned_df$direction == "up"]),
                                               no_ctm_binned_df$temperature[no_ctm_binned_df$direction == "down"][1]),
                               salinity = c(no_ctm_binned_df$salinity[no_ctm_binned_df$direction == "down"],
                                            rev(no_ctm_binned_df$salinity[no_ctm_binned_df$direction == "up"]),
                                            no_ctm_binned_df$salinity[no_ctm_binned_df$direction == "down"][1]),
                               direction = c(rep("down", nrow(no_ctm_binned_df)/2),
                                             rep("up", nrow(no_ctm_binned_df)/2),
                                             "down")) |>
                      sf::st_as_sf(coords = c("temperature", "salinity")) |>
                      dplyr::group_by(ID = 1) |>
                      summarise(do_union = FALSE) |>
                      sf::st_cast(to = "POLYGON") |>
                      sf::st_area(),
                    ctm_adjust_tsarea(alpha = 0.04,
                                      tau = 8,
                                      f_n = 0.25,
                                      down_temperature = no_ctm_df$temperature[no_ctm_df$direction == "down"],
                                      down_pressure = no_ctm_df$pressure[no_ctm_df$direction == "down"],
                                      down_conductivity = no_ctm_df$conductivity[no_ctm_df$direction == "down"],
                                      down_flag = no_ctm_df$flag[no_ctm_df$direction == "down"],
                                      up_temperature = no_ctm_df$temperature[no_ctm_df$direction == "up"],
                                      up_pressure = no_ctm_df$pressure[no_ctm_df$direction == "up"],
                                      up_conductivity = no_ctm_df$conductivity[no_ctm_df$direction == "up"],
                                      up_flag = no_ctm_df$flag[no_ctm_df$direction == "up"],
                                      obj_fn = "area"),
                    ctm_adjust_tsarea(alpha = optim_results@coef['alpha'],
                                      tau = optim_results@coef['tau'],
                                      f_n = 0.25,
                                      down_temperature = no_ctm_df$temperature[no_ctm_df$direction == "down"],
                                      down_pressure = no_ctm_df$pressure[no_ctm_df$direction == "down"],
                                      down_conductivity = no_ctm_df$conductivity[no_ctm_df$direction == "down"],
                                      down_flag = no_ctm_df$flag[no_ctm_df$direction == "down"],
                                      up_temperature = no_ctm_df$temperature[no_ctm_df$direction == "up"],
                                      up_pressure = no_ctm_df$pressure[no_ctm_df$direction == "up"],
                                      up_conductivity = no_ctm_df$conductivity[no_ctm_df$direction == "up"],
                                      up_flag = no_ctm_df$flag[no_ctm_df$direction == "up"],
                                      obj_fn = "area")),
           alpha = c(NA, 0.04, optim_results@coef['alpha']),
           tau = c(NA, 8, optim_results@coef['tau']))

print(ts_area_df)
```

```{r, include=TRUE, include=FALSE, warning=FALSE, results ='hide'}
optim_ctm_df <- ctm_adjust_tsarea(alpha = optim_results@coef['alpha'],
                             tau = optim_results@coef['tau'],
                             f_n = 0.25,
                             down_temperature = no_ctm_df$temperature[no_ctm_df$direction == "down"],
                             down_pressure = no_ctm_df$pressure[no_ctm_df$direction == "down"],
                             down_conductivity = no_ctm_df$conductivity[no_ctm_df$direction == "down"],
                             down_flag = no_ctm_df$flag[no_ctm_df$direction == "down"],
                             up_temperature = no_ctm_df$temperature[no_ctm_df$direction == "up"],
                             up_pressure = no_ctm_df$pressure[no_ctm_df$direction == "up"],
                             up_conductivity = no_ctm_df$conductivity[no_ctm_df$direction == "up"],
                             up_flag = no_ctm_df$flag[no_ctm_df$direction == "up"],
                             obj_fn = "none") |>
  dplyr::filter(pressure_bin > 4)

default_ctm_df <- ctm_adjust_tsarea(alpha = 0.04,
                             tau = 8,
                             f_n = 0.25,
                             down_temperature = no_ctm_df$temperature[no_ctm_df$direction == "down"],
                             down_pressure = no_ctm_df$pressure[no_ctm_df$direction == "down"],
                             down_conductivity = no_ctm_df$conductivity[no_ctm_df$direction == "down"],
                             down_flag = no_ctm_df$flag[no_ctm_df$direction == "down"],
                             up_temperature = no_ctm_df$temperature[no_ctm_df$direction == "up"],
                             up_pressure = no_ctm_df$pressure[no_ctm_df$direction == "up"],
                             up_conductivity = no_ctm_df$conductivity[no_ctm_df$direction == "up"],
                             up_flag = no_ctm_df$flag[no_ctm_df$direction == "up"],
                             obj_fn = "none") |>
  dplyr::filter(pressure_bin > 4)

ctm_corr_df <- dplyr::bind_rows(
  dplyr::bind_rows(
    default_ctm_df |>
      dplyr::select(pressure_bin, temperature_down, salinity_down) |>
      dplyr::rename(temperature = temperature_down,
                    salinity = salinity_down) |>
      dplyr::select(salinity, pressure_bin) |>
      dplyr::mutate(direction = "down"),
    default_ctm_df |>
      dplyr::select(pressure_bin, temperature_up, salinity_up) |>
      dplyr::rename(temperature = temperature_up,
                    salinity = salinity_up) |>
      dplyr::select(salinity, pressure_bin) |>
      dplyr::mutate(direction = "up")) |>
    dplyr::mutate(ctm = "Default CTM"),
  dplyr::bind_rows(
    optim_ctm_df |>
      dplyr::select(pressure_bin, temperature_down, salinity_down) |>
      dplyr::rename(temperature = temperature_down,
                    salinity = salinity_down) |>
      dplyr::select(salinity, pressure_bin) |>
      dplyr::mutate(direction = "down"),
    optim_ctm_df |>
      dplyr::select(pressure_bin, temperature_up, salinity_up) |>
      dplyr::rename(temperature = temperature_up,
                    salinity = salinity_up) |>
      dplyr::select(salinity, pressure_bin) |>
      dplyr::mutate(direction = "up")) |>
    dplyr::mutate(ctm = "Optimized CTM"),
  no_ctm_df |>
    dplyr::select(pressure, salinity, direction, flag) |>
    dplyr::filter(pressure > 4, !is.na(flag)) |>
    dplyr::mutate(pressure_bin = round(pressure)) |>
    dplyr::group_by(direction, pressure_bin) |>
    dplyr::summarise(salinity = mean(salinity)) |>
    dplyr::mutate(ctm = "No CTM"))

ts_area_polygon <- dplyr::bind_rows(data.frame(temperature = c(default_ctm_df$temperature_down,
                                                    rev(default_ctm_df$temperature_up),
                                                    default_ctm_df$temperature_down[1]),
                              salinity = c(default_ctm_df$salinity_down,
                                                 rev(default_ctm_df$salinity_up),
                                                 default_ctm_df$salinity_down[1]),
                              direction = c(rep("down", nrow(default_ctm_df)),
                                            rep("up", nrow(default_ctm_df)),
                                            "down")) |>
  dplyr::mutate(ctm = "Default CTM"),
data.frame(temperature = c(no_ctm_binned_df$temperature[no_ctm_binned_df$direction == "down"],
                           rev(no_ctm_binned_df$temperature[no_ctm_binned_df$direction == "up"]),
                           no_ctm_binned_df$temperature[no_ctm_binned_df$direction == "down"][1]),
           salinity = c(no_ctm_binned_df$salinity[no_ctm_binned_df$direction == "down"],
                        rev(no_ctm_binned_df$salinity[no_ctm_binned_df$direction == "up"]),
                        no_ctm_binned_df$salinity[no_ctm_binned_df$direction == "down"][1]),
           direction = c(rep("down", nrow(no_ctm_binned_df)/2),
                         rep("up", nrow(no_ctm_binned_df)/2),
                         "down")) |>
  dplyr::mutate(ctm = "No CTM"),
data.frame(temperature = c(optim_ctm_df$temperature_down,
                           rev(optim_ctm_df$temperature_up),
                           optim_ctm_df$temperature_down[1]),
           salinity = c(optim_ctm_df$salinity_down,
                        rev(optim_ctm_df$salinity_up),
                        optim_ctm_df$salinity_down[1]),
           direction = c(rep("down", nrow(optim_ctm_df)),
                         rep("up", nrow(optim_ctm_df)),
                         "down")) |>
  dplyr::mutate(ctm = "Optimized CTM"))

```

Optimized CTM has a substantially smaller area between T-S curves than both No CTM and Default CTM. Meanwhile, corrections using the the typical values (Default CTM) result in a larger T-S area than the profile without a correction (No CTM). The difference in area between curves is evident in T-S plots (Figure 2) and salinity profiles (Figure 3) for No CTM, Default CTM, and Optimized CTM. In the Optimized CTM profile, there are no salinity spikes around the pycnocline in the upcast and the downcast spike is smaller than in the other profiles.

```{r, include=FALSE, echo=TRUE}
ggplot() +
  geom_polygon(data = ts_area_polygon,
               aes(x = temperature,
                   y = salinity),
               fill = "grey70",
               color = "black") +
  geom_point(data = ts_area_polygon,
             aes(x = temperature,
                 y = salinity,
                 color = direction)) +
  geom_text(data = ts_area_df,
             aes(x = 5, y = 32.39, label = paste0("Area = ", formatC(area, format = "e", digits = 2), "\nalpha = ", round(alpha, 2), "\ntau = ", round(tau, 1))),
            hjust = 0) +
  scale_color_colorblind(name = "Direction") +
  scale_y_continuous(name = "Salinity, PSS-78") +
  scale_x_continuous(name = expression(Temperature~(degree*C))) +
  facet_wrap(~factor(ctm, levels = c("No CTM", "Default CTM", "Optimized CTM"))) +
  theme_bw() +
  theme(legend.position = "bottom")
```
<i>Figure 2. Temperature-salinity profiles and area between upcast and downcast profiles. Panels show areas between T-S curves without cell thermal mass correction (No CTM), with cell thermal mass correction using default parameter values (Default CTM), and with cell thermal mass correction using optimized parameters (Optimized CTM).</i>

```{r, include=FALSE, echo=TRUE}
ggplot() +
  geom_path(data = ctm_corr_df,
            aes(x = salinity,
                y = pressure_bin,
                color = direction),
            size = rel(1.2)) +
  scale_color_colorblind(name = "Direction") +
  scale_x_continuous(name = "Salinity, PSS-78") +
  scale_y_reverse(name = "Pressure (db)") +
  facet_wrap(~factor(ctm, levels = c("No CTM", "Default CTM", "Optimized CTM")),
             scales = "free_x") +
  theme_bw() +
  theme(legend.position = "bottom")
```
<i>Figure 3. Upcast and downcast profiles of salinity without cell thermal mass correction (No CTM), with cell thermal mass correction using default parameter values (Default CTM), and with cell thermal mass correction using optimized parameters (Optimized CTM).</i>


## References

Garau, B., S. Ruiz, W. G. Zhang, A. Pascual, E. Heslop, J. Kerfoot, and J. Tintoré, 2011: Thermal lag correction on slocum CTD glider data. J. Atmos. Ocean. Technol., 28, 1065–1071, [https://doi.org/10.1175/JTECH-D-10-05030.1](https://doi.org/10.1175/JTECH-D-10-05030.1)

Lueck, R. G., and J. J. Picklo, 1990: Thermal Inertia of Conductivity Cells: Observations with a Sea-Bird Cell. J. Atmos. Ocean. Technol., 7, 756–768, [https://doi.org/10.1175/1520-0426(1990)007<0756:TIOCCO>2.0.CO;2](https://doi.org/10.1175/1520-0426(1990)007<0756:TIOCCO>2.0.CO;2)

Martini, K. I., D. J. Murphy, R. W. Schmitt, and N. G. Larson, 2019: Corrections for pumped SBE 41CP CTDs determined from stratified tank experiments. J. Atmos. Ocean. Technol., 36, 733–744, [https://doi.org/10.1175/JTECH-D-18-0050.1](https://doi.org/10.1175/JTECH-D-18-0050.1)

Mensah, V., M. Le Menn, and Y. Morel, 2009: Thermal mass correction for the evaluation of salinity. J. Atmos. Ocean. Technol., 26, 665–672, [https://doi.org/10.1175/2008JTECHO612.1](https://doi.org/10.1175/2008JTECHO612.1)
